`{peacesciencer}`: An R Package for Quantitative Peace Science Research

### 2021-11-09 user-guide/ms.Rmd
upload/update, ms will be a work in progress.

---
title: "`{peacesciencer}`: An R Package for Quantitative Peace Science Research"
thanks: "Replication files are available on the author's Github account (https://github.com/svmiller/peacesciencer). **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: steven.v.miller@gmail.com."
author:
- name: Steven V. Miller
  affiliation: Clemson University
bibliography: "`r paste0(Sys.getenv('HOME'),'/Dropbox/master.bib')`" # This would be more general
# bibliography: inst/refs.bib
abstract: "This article introduces `{peacesciencer}`, an R package that contains a litany of tools for creating data of widespread interest to the peace science community. The package is cross-platform, assuming only a somewhat recent installation of the R programming language with some of the enhanced functionality of the broadly popular `{tidvyerse}` packages. Peace science researchers can use this package to greatly reduce the time needed to perfectly recreate dyad-year and state-year data from scratch and merge in common indicators included in almost every dyad-year or state-year analysis (e.g. democracy data, contiguity data). The software is freely available on CRAN and maintains an active website documenting its features at `r params$ps_website`."
keywords: "software, statistical analysis, peace science, reproducibility"
geometry: margin=1in
csl: inst/jpr.csl
mainfont: cochineal
#sansfont: Linux Biolinum O 
fontsize: 11pt
endnote: no
# pandocparas: TRUE
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: ""
  removetitleabstract: ""
  ps_website: ""
  ps_manuals: ""
  ps_github: ""
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
#always_allow_html: TRUE
header-includes:
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{csquotes}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
#  - \let\BeginKnitrBlock\begin \let\EndKnitrBlock\end
#  - \usepackage{colortbl}
#  - \arrayrulecolor{black}
#  - \usepackage{threeparttable}
#  - \usepackage{threeparttablex}
#  - \midrule\relax
---

@@

This data feature tackles a recurring problem for researchers in the peace science community. True research reproducibility is achieved creating original dyad-year or state-year data from scratch, though no published guide exists that informs researchers how to do this on their own. Instead, researchers may end up reusing old code generated in past studies, leaving them to spend time and energy adjusting the sample of states, the temporal domain, and doing whatever additional troubleshooting may arise from this practice. Researchers may additionally spend too much time reproducing old code for standard information that goes into any dyad-year or state-year analysis---like contiguity relationships and democracy---and have to do additional troubleshooting for how these various data sources treat missing data or treat state codes in a manner inconsistent with the more accessible Correlates of War or Gleditsch-Ward state codes. This is all compounded by changes in technology that treat the creation of the data and the analysis of data as a continuous process in which the contemporary quantitative political scientist is increasingly becoming a computer programmer as well. Graduate students, in particular, face unique challenges associated with these developments. Graduate students beginning their careers in peace science must learn how scholarship informs data in peace science and how data inform scholarship at the same time they are needing to learn quantitative methods in a chosen software package.

`{peacesciencer}` is designed to address these problems. Built around the free and open source R programming language, `{peacesciencer}` contains a suite of data and functions for creating data of interest to researchers. Researchers can use `{peacesciencer}` to create dyad-year, state-year, and even state-day data from scratch. Afterward, they can add a variety of standard information (e.g. contiguity, alliances, major power status, GDP per capita estimates, capability estimates, and more) to these data with a simple command. This is a considerable time-saver since, in the absence of it, researchers would have to more meticulously code and transform the raw data to conform to the kind of data they want. `{peacesciencer}` comes with some data innovations as well, including a comprehensive data set on democracy by year, an original data set on capitals and capital transitions, and a function to create peace years between ongoing dyad-year or state-year conflicts. All are done with the maximum possible transparency. The project is available for public view of Github (`r params$ps_github`). The `data-raw` directory on the project's Github contains information and comments about how every data set was created and the `R` directory informs the user how each function works. The function manuals (`r params$ps_manual`) contain additional comments about what each function returns and, in appropriate cases, why it is doing what it is doing. Thus, `{peacesciencer}` not only assists a peace scientist with their research, but it does so in a manner that best conforms to the  Data Access and Research Transparency Initiative (DA-RT) initiative across all political science.

@@

For one, researchers invest too much time in the construction of a data set that faithfully captures the unit of analysis. Assume a researcher wants an original data set on all directed dyad-years for Correlates of War states for an analysis of inter-state conflict. How might one do that? The answer has never been immediately obvious. No published guide exists that shows a researcher how to create these data themselves from scratch, which is one reason why software bundles like EUGene and NewGene are attractive to researchers who primarily care about the substance of their research question. After all, EUGene's primary impetus was allowing researchers to create data sets for replication of previous studies, producing a host of data types (e.g. dyad-year, state-year, dispute-year) along the way that users can amend as they saw fit. `{peacesciencer}` is primarily born from this question about how to create dyad-year and state-year data from scratch. The underlying code that produces these data types is available online and `{peacesciencer}` converts these lines of code into simple functions for the ease of the researcher.

Second, researchers also invest too much time in retracing steps for peace science analyses for new projects. Assume  a researcher who had finished a state-year analysis on the correlates of civil conflict onset a couple years ago and wants to start a new project that analyzes the same outcome from a different angle (or perhaps using newer data). Under these conditions, a researcher will have to find where they stored that replication code and copy-paste it into a new directory for the new project. They may then have to change the name of some files, change some code to account for potentially new column names in the newer data, and troubleshoot instances where their old code does not perform as it once did. At its worst, this process may lead to some errors by the researcher. At best, this is tedium that spends the researcher's time they would rather invest in analyzing the data. The lion's share of `{peacesciencer}`'s functionality is both creating the units of analysis for the researcher and merging in different forms of data in wide use in the peace science community so that the researcher can spend less of their time on tedium.

@@

Fifth, statistical software should be kind to graduate students in peace science with the aforementioned points in mind. Graduate students just beginning their career in peace science scholarship are tasked to learn the substantive material, the characteristics of the underlying data in wide use in the community, and the quantitative methods and statistical software available to them, effectively all at the same time. A student interested in civil conflict scholarship will need to learn the UCDP-PRIO armed conflict data in some detail at the same time they learn about the theoretical arguments that informed the creation of the data, likely at the same time they take quantitative methods courses that show them how to load and explore data in a chosen statistical software. At the same time that creating data is increasingly blurred with coding and processing data, students are simultaneously learning to explore data for themselves as they read about the analyses of the data. `{peacesciencer}` aims to make this process easier for graduate students using the features of the R programming language. Graduate students can use `{peacesciencer}` to learn about peace science data and computer programming in R at the same time.

@@

`{peacesciencer}` has a simple "pipe"-based workflow. The top of the "pipe" is the first function to create the data. These options include `create_dyadyears()`, `create_stateyears()`, and `create_statedays()`. All options include two arguments: `system` and `mry`. `system` takes one of two values---"cow" or "gw"---for what state system membership data the user wants. If `system` is not specified, the function defaults to Correlates of War system membership data given the prominence of the Correlates of War project in the conflict data ecosystem. `mry`---short for "most recent year"---determines whether or not the function extends the membership data to the most recently concluded calendar year. The function defaults to `TRUE` in the absence of a user-specified override, meaning the data returned will include all relevant state-year data from 1816 to (as of writing) 2020 even though the Correlates of War state system data were last updated at the end of 2016 (and the Gleditsch-Ward system were last updated at the end of 2017). `create_dyadyears()` has a third argument---`directed`---that determines whether the user wants directed dyad-year data or non-directed dyad-year data. The function defaults to `TRUE`, returning directed dyad-year data.

The following example creates a full state-year data frame of Gleditsch-Ward states from 1816 to 2020.

The below example would create a full non-directed dyad-year data frame of Correlates of War states from 1816 to 2016, the last year of record in the Correlates of War state system membership data.

```{r, echo=T, eval=F}
create_dyadyears(directed = FALSE, mry = FALSE)
```

This example would create a full state-day data frame from the Gleditsch-Ward system for all states from Jan. 1, 1816 to Dec. 31, 2020. 
```{r, echo=T, eval=F}
create_statedays(system = 'gw')
```

There are not too many data sets in the peace science community with daily data, certainly dating to the 19th century, but `{peacesciencer}` can greatly reduce the costs of creating the kind of data the user may want.

## Add Standard Information to Dyad-Year or State-Year Data

The next step in the `{peacesciencer}` "pipe"-based workflow will also be among the biggest time-savers for researchers using it. Researchers who want to understand, say, the correlates of inter-state conflict onset almost always do the following. They will add the dependent variables from an inter-state conflict data set, like the Gibler-Miller-Little (GML) dispute data set [@gibleretal2016amid]. They will add several indicators to the dyad-year data that are well-traveled in conflict scholarship, including indicators of things like joint alliance contracts and the contiguity relationship of both states in the dyad. `{peacesciencer}` has great coverage for a lot of these data and can even allow almost full-scale replications with the data it offers.

For example, assume a researcher wanted to replicate something analogous to Bremer's [-@bremer1992ddc] famous "dangerous dyads" analysis about the causes of conflict onset. `{peacesciencer}` can do something like this in the following pipe-based workflow. First, `create_dyadyears(directed = FALSE, mry = FALSE)` will create a non-directed dyad-year data frame for all Correlates of War states from 1816 to 2016.[^after2010]

[^after2010]: The conflict data in @gibleretal2016amid do not extend past 2010, but asking for dyad-years from 2010-2016 will just create harmless `NA`s that will not appear in any analysis.

After creating the base data they want, the user can add data to it with the "pipe" operator. The `{tidyverse}` represents this pipe operator as `%>%`. Ending each line with this pipe passes forward an object (here: data) into the next function on the next line. Put in other words, `create_dyadyears(directed = FALSE, mry = FALSE) %>% filter_prd()` is a command in `{tidyverse}` and `{peacesciencer}` to create all non-directed dyad-years from 1816 to 2016 and then subset those data to only the politically relevant dyad-years. Because politically relevant dyads depend on information about the major power status for both states as well as their contiguity relationship [@lemkereed2001rprd], the function adds in those data (i.e. quietly executing `add_contiguity()` and `add_cow_majors()`) before subsetting the data to just the dyads that have some kind of contiguity relationship and/or have a major power in it.

```{r, echo=T, eval=F}
create_dyadyears(directed = FALSE, mry = FALSE) %>%
  filter_prd() %>%
  add_gml_mids(keep = NULL) %>%
  add_peace_years() %>%
  add_nmc() %>%
  add_democracy() %>%
  add_cow_alliance() %>%
  add_sdp_gdp() -> Data
```

Next, the pipe-based workflow continues with `add_gml_mids(keep = NULL)`, which will add information about whether there was an ongoing dispute in the dyad (and whether it was a dispute onset) from the GML dispute data.[^keepnull] `add_peace_years()` will calculate peace years between ongoing disputes in these politically relevant non-directed dyads.[^onpy] `add_nmc()` will add information from the Correlates of War National Military Capabilities data, prominently the CINC scores that conflict scholars routinely use for assessing hypotheses of power preponderance and power parity. `add_democracy()` will add information about the level of democracy for both states in the dyad using three different indicators. Researchers can select which one they prefer from these indicators. `add_cow_alliance()` will add information about whether there was an alliance in the dyad-year by way of whether there was an agreement that included a defense pledge, a neutrality pledge, a non-aggression pledge, or an entente pledge. Finally, `add_sdp_gdp()` will add information about GDP, GDP per capita, and the surplus domestic product of both states in the dyad from the simulations done by @andersetal2020bbgb.

[^keepnull]: `keep = NULL` is optional and removes supporting information about dyadic dispute-participants (e.g. dispute-level hostility, dispute-level outcome) in order to reduce the size of the data set.

@@

Whereas `add_sdp_gdp()` is the last command in the pipe-based workflow, the `{peacesciencer}` call ends by assigning to an object called `Data`. This type of assignment is done with the "right hand" assignment operator (i.e. `->`). If the user wants to move these data into Stata for analysis, they can save it to their current working directory with a command like `haven::write_dta(Data, "my-data.dta")` and import it into Stata when they are done.

`{peacesciencer}` has pre-processed, cleaned, recoded, and merged the desired data that has greatly reduced the time and energy a researcher might otherwise spend doing something like hard-coding `-9s` in these data to be `NA` in the National Material Capabilities data. There is only some slight data work to create the desired indicators for a statistical model of conflict onset. Here, briefly, these functions lean on `{tidyverse}` "verbs" and base R terminology to do things like creating a dummy variable for whether the dyad was land-contiguous, whether there was a major power on either side of the non-directed dyad, creating some "weak-link" indicators of militarization, relative power in the dyad, level of democracy in the dyad (using the @marquez2016qme for extending the Unified Democracy Scores data), and the GDP per capita in the dyad.

```{r, echo=T, eval=F}
Data %>%
  mutate(landcontig = ifelse(conttype == 1, 1, 0)) %>%
  mutate(cowmajdyad = ifelse(cowmaj1 == 1 | cowmaj2 == 1, 1, 0)) %>%
  # Create estimate of militarization as milper/tpop
  # Then make a weak-link
  mutate(milit1 = milper1/tpop1,
         milit2 = milper2/tpop2,
         minmilit = ifelse(milit1 > milit2,
                           milit2, milit1)) %>%
  # create CINC proportion (lower over higher)
  mutate(cincprop = ifelse(cinc1 > cinc2,
                           cinc2/cinc1, cinc1/cinc2)) %>%
  # create weak-link specification using Quick UDS data
  mutate(mindemest = ifelse(xm_qudsest1 > xm_qudsest2,
                            xm_qudsest2, xm_qudsest1)) %>%
  # Create "weak-link" measure of jointly advanced economies
  mutate(minwbgdppc = ifelse(wbgdppc2011est1 > wbgdppc2011est2,
                             wbgdppc2011est2, wbgdppc2011est1)) -> Data
```

Table \@ref(tab:tab-dd) is a formatted version of the results of a logistic regression model of conflict onset using these "dangerous dyads" indicators. Users typically do not end their analysis here---often looking for new predictors of conflict onset with these covariates in mind---but `{peacesciencer}` greatly reduces the time and energy researchers must invest into cleaning and processing data for analysis.

@@

`{peacesciencer}` is also capable of creating data for replications of standard civil conflict analyses. Suppose a researcher wants to create a state-year data frame to conduct an analysis of civil conflict onset analogous to Fearon and Laitin's [-@fearonlaitin2003eicw] well-cited analysis of civil conflict onset, but using UCDP conflict data and the Gleditsch-Ward state system for creating the appropriate universe of state-years. The pipe-based workflow will start with `create_stateyears(system = 'gw')`, creating the full universe of Gleditsch-Ward state years from 1816 to 2020. Whereas the UCDP data included in `{peacesciencer}` are from 1946 to 2019 (v. 20.1), we can subset ("filter") the data to just those years with `filter(year %in% c(1946, 2019))`. Next, we can use the `add_ucdp_acd()` function to return information about ongoing UCDP conflicts and onsets for these states. `add_ucdp_acd()` takes three arguments: `type`, `issue`, and `only_wars`. `type` is an optional argument for the type of armed conflicts for which the researcher wants information. Options include "extrasystemic", "interstate", "intrastate", and "II" (short for "internationalized intrastate"). If no `type` is specified, the function returns information about ongoing disputes and onsets for all states for all tyeps of conflict. If the user wants information about multiple types of conflict--say: intra-state wars and internationalized intra-state wars---they can specify that as a character vector (e.g. `type = c("intrastate", "II")`). `issue` is another optional argument for what issue types of conflicts the user wants beyond the type of armed conflict. Options include "territory", "government", and "both". If no `issue` is specified, the function returns information for all conflicts irregarding the particular issue. `only_wars` is an argument that subsets the data to just those with the intensity levels of "war" when `only_wars = TRUE`. The argument defaults to `FALSE`, returning information about conflicts with at least 25 deaths in addition to the conflicts with more than 1,000 deaths. In this application, `add_ucdp_acd(type = "intrastate", only_wars = FALSE))` returns state-year information about ongoing intra-state conflicts over any issue and at either of UCDP's severity thresholds.[^adducdpacdalso]

@@

```{r, echo=T, eval=F}
create_stateyears(system = 'gw') %>%
  filter(year %in% c(1946:2019)) %>%
  add_ucdp_acd(type="intrastate", only_wars = FALSE) %>%
  add_peace_years() %>%
  add_democracy() %>%
  add_creg_fractionalization() %>%
  add_sdp_gdp() %>%
  add_rugged_terrain() -> Data
```

@@

```{r, echo=T, eval=F}
create_stateyears(system = 'gw') %>%
  filter(year %in% c(1946:2019)) %>%
  add_ucdp_acd(type="intrastate", only_wars = TRUE) %>%
  add_peace_years() %>%
  rename_at(vars(ucdpongoing:ucdpspell), ~paste0("war_", .)) %>%
  left_join(Data, .) -> Data
```

@@

`{peacesciencer}` is already more than capable of creating the kind of data in high demand in peace science. It can create dyad-year and state-year data. It is also generalizable to the dispute data included in the package, allowing for merging into *dispute-year* data as well. This feature showed how it can effectively approximate two types of analyses in wide use in the peace science community. Surely researchers can and will add more information to these simple analyses after using `{peacesciencer}`, but the package already does a lot of the tedious work for researchers. It also does this in a maximally transparent way that conforms well to the DA-RT initiative across all political science.

>>>>

---
title: "`{peacesciencer}`: An R Package for Quantitative Peace Science Research"
thanks: "Replication files are available on the author's Github account (https://github.com/svmiller/peacesciencer). **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: steven.v.miller@gmail.com."
author:
- name: Steven V. Miller
  affiliation: Clemson University
bibliography: "`r paste0(Sys.getenv('HOME'),'/Dropbox/master.bib')`" # This would be more general
# bibliography: inst/refs.bib
abstract: "This article introduces `{peacesciencer}`, an R package that contains a litany of tools for creating data of widespread interest to the peace science community. The package is cross-platform, assuming only a somewhat recent installation of the R programming language with some of the enhanced functionality of the broadly popular `{tidvyerse}` packages. Peace science researchers can use this package to greatly reduce the time needed to perfectly recreate dyad-year and state-year data from scratch and merge in common indicators included in almost every dyad-year, leader-year, leader-dyad-year, or state-year analysis (e.g. democracy data, contiguity data). The software is freely available on CRAN and maintains an active website documenting its features at `r params$ps_website`."
keywords: "software, statistical analysis, peace science, reproducibility"
geometry: margin=1in
csl: inst/jpr.csl
mainfont: cochineal
#sansfont: Linux Biolinum O 
fontsize: 11pt
endnote: no
# pandocparas: TRUE
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: ""
  removetitleabstract: ""
  ps_website: ""
  ps_manuals: ""
  ps_github: ""
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
#always_allow_html: TRUE
header-includes:
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{csquotes}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
#  - \let\BeginKnitrBlock\begin \let\EndKnitrBlock\end
#  - \usepackage{colortbl}
#  - \arrayrulecolor{black}
#  - \usepackage{threeparttable}
#  - \usepackage{threeparttablex}
#  - \midrule\relax
---

@@

This data feature tackles a recurring problem for researchers in the peace science community. True research reproducibility is best achieved creating data from scratch, though no published guide exists that informs researchers how to do this on their own. Instead, researchers may end up reusing old code generated in past studies, leaving them to spend time and energy adjusting the sample of states, the temporal domain, and doing whatever additional troubleshooting may arise from this practice. Researchers may additionally spend too much time reproducing old code for standard information that goes into any dyadic or monadic analysis---like contiguity relationships and democracy---and have to do additional troubleshooting for how these various data sources treat missing data or treat state codes in a manner inconsistent with the more accessible Correlates of War or Gleditsch-Ward state codes. This is all compounded by changes in technology that treat the creation of the data and the analysis of data as a continuous process in which the contemporary quantitative political scientist is increasingly becoming a computer programmer as well. Graduate students, in particular, face unique challenges associated with these developments. Graduate students beginning their careers in peace science must learn how scholarship informs data in peace science and how data inform scholarship at the same time they are needing to learn quantitative methods in a chosen software package.

`{peacesciencer}` is designed to address these problems. Built around the free and open source R programming language, `{peacesciencer}` contains a suite of data and functions for creating data of interest to researchers. Researchers can use `{peacesciencer}` to create dyad-year, leader-year, leader-dyad-year, and state-year data (among some others) from scratch. Afterward, they can add a variety of standard information (e.g. contiguity, alliances, major power status, GDP per capita estimates, capability estimates, and more) to these data with a simple command. This is a considerable time-saver since, in the absence of it, researchers would have to more meticulously code and transform the raw data to conform to the kind of data they want. `{peacesciencer}` comes with some data innovations as well, including a comprehensive data set on democracy by year, an original data set on capitals and capital transitions, and a function to create peace years between ongoing conflicts. All are done with the maximum possible transparency. The project is available for public view of Github (`r params$ps_github`). The `data-raw` directory on the project's Github contains information and comments about how every data set was created and the `R` directory informs the user how each function works. The function manuals (`r params$ps_manual`) contain additional comments about what each function returns and, in appropriate cases, why it is doing what it is doing. Thus, `{peacesciencer}` not only assists a peace scientist with their research, but it does so in a manner that best conforms to the  Data Access and Research Transparency Initiative (DA-RT) initiative across all political science.

@@

For one, researchers invest too much time in the construction of a data set that faithfully captures the unit of analysis. Assume a researcher wants an original data set on all directed dyad-years for Correlates of War states for an analysis of inter-state conflict. How might one do that? The answer has never been immediately obvious. No published guide exists that shows a researcher how to create these data themselves from scratch, which is one reason why software bundles like EUGene and NewGene are attractive to researchers who primarily care about the substance of their research question. After all, EUGene's main value---if not its primary impetus---was allowing researchers to create data sets for replication of previous studies, producing a host of data types (e.g. dyad-year, state-year, dispute-year) along the way that users can amend as they saw fit. `{peacesciencer}` is primarily born from this question about how to create these data from scratch. The underlying code that produces these data types is available online and `{peacesciencer}` converts these lines of code into simple functions for the ease of the researcher.

Second, researchers also invest too much time in retracing steps for peace science analyses for new projects. Assume  a researcher finished a state-year analysis on the correlates of civil conflict onset a few years ago and wants to start a new project that analyzes the same outcome from a different angle (or perhaps using newer data). Under these conditions, a researcher will have to find where they stored that replication code and copy-paste it into a new directory for the new project. They may then have to change the name of some files, change some code to account for potentially new column names in the newer data, and troubleshoot instances where their old code does not perform as it once did. At its worst, this process may lead to some errors by the researcher. At best, this is tedium that spends the researcher's time they would rather invest in analyzing the data. The lion's share of `{peacesciencer}`'s functionality is both creating the units of analysis for the researcher and merging in different forms of data in wide use in the peace science community so that the researcher can spend less of their time on tedium.

@@

<!-- Fifth, statistical software should be kind to graduate students in peace science with the aforementioned points in mind. Graduate students just beginning their career in peace science scholarship are tasked to learn the substantive material, the characteristics of the underlying data in wide use in the community, and the quantitative methods and statistical software available to them, effectively all at the same time. A student interested in civil conflict scholarship will need to learn the UCDP-PRIO armed conflict data in some detail at the same time they learn about the theoretical arguments that informed the creation of the data, likely at the same time they take quantitative methods courses that show them how to load and explore data in a chosen statistical software. At the same time that creating data is increasingly blurred with coding and processing data, students are simultaneously learning to explore data for themselves as they read about the analyses of the data. `{peacesciencer}` aims to make this process easier for graduate students using the features of the R programming language. Graduate students can use `{peacesciencer}` to learn about peace science data and computer programming in R at the same time.  -->

@@

Researchers using `{peacesciencer}` to create data for their research project should start with one of these "create" functions. Whether state-year, dyad-year, or a leader-level analysis, these functions will create the full universe of cases of interest to a researcher for the unit of analysis that serves as the core of their project.

## Adding Information to Data Created in `{peacesciencer}`

After creating the base data of interest to their project, researchers can begin to add information they want with the suite of functions outlined in Table \@ref(tab:tab-addfunctions). For example, researcher interested in a dyad-year analysis of inter-state disputes can create a non-directed dyad-year data set from 1816 to 2010 in `{peacesciencer}` with the `create_dyadyears()` function, one of the aforementioned "create" function. The following would create the base data of interest to the user (i.e. all non-directed dyad-years from 1816 to 2010).

```{r}
create_dyadyears(directed = FALSE, subset_years = c(1816:2010))
```

Adding information to this data frame is a simple matter of joining a series of functions together in a "pipe." The "pipe"---represented as `%>%` in the code below---is an operator built into `{tidyverse}` that allows (`R`) users to pass forward expressions or functions. These pipes are common in the programming world and, as the code below will show, have the benefit of changing code in a way that is more intuitive and easier to both read and write for the user. While these functions can be modified to work without `{tidyverse}` installed and loaded into the session, the user should find their experience with `{peacesciencer}` improved by this important package.

For example, assume the researcher wants just all politically relevant, non-directed dyad-years, where political relevance is traditionally understood as a dyadic relationship involving some form of a contiguity relationship or a major power [@lemkereed2001rprd]. `create_dyadyears(directed = FALSE, subset_years = c(1816:2010))` created the full universe of non-directed dyad-years from 1816 to 2010, though this full universe includes "irrelevant" dyads like Nigeria-Mongolia and Canada-Estonia. Reducing the data to just politically relevand dyads is simple in `{peacesciencer}` and `{tidyverse}`. Users first create the data they want (here: `create_dyadyears(directed = FALSE, subset_years = c(1816:2010))`), follow it with the pipe operator (`%>%`), and then add another function from `{peacesciencer}` (here: `filter_prd()`, which also quietly executes `add_contiguity()` and `add_cow_majors()`).

```{r}
# create base data, and pipe (%>%) to next function
create_dyadyears(directed = FALSE, subset_years = c(1816:2010)) %>%
  # subset data to politically relevant dyads (PRDs)
  filter_prd()
```

Here, the user has created all non-directed dyad-years from 1816 to 2010 and then subset the data to just those with a major power or with some kind of contiguity relationship.

Users will find that the ease of the "pipe" will allow them greater agency in creating the full data set they may want for an analysis. Indeed, the pipe has the effect of forming something analogous to a drop-down menu, in which the user can "select" additional data they may want simply by specifying the function in `{peacesciencer}` that creates the data for them. For example, a researcher can follow `filter_prd()` with another pipe and communicate they want information about ongoing conflicts and conflict onsets from the Gibler-Miller-Little (GML) dispute data set [@gibleretal2016amid]. Following `filter_prd()` with `add_gml_mids(keep = NULL)` will add information about ongoing conflicts and onsets in a given dyad-year.[^keepnull]

[^keepnull]: `keep = NULL` in this context will focus the information returned to just the information about ongoing conflicts and onsets, discarding potentially unwanted columns about things like Side A, hostility levels, and other information included in the data set. Type `?add_gml_mids()` for more information.

```{r}
# create base data, and pipe (%>%) to next function
create_dyadyears(directed = FALSE, subset_years = c(1816:2010)) %>%
  # subset data to politically relevant dyads (PRDs), pipe to next function
  filter_prd() %>%
  # add conflict information from GML-MID data,
  add_gml_mids(keep = NULL) 
```

Users can also calculate peace-years for these conflicts with `add_peace_years()` by using the pipe to pass forward the data set and applying the `add_peace_years()` function to it.[^onpy]

@@

```{r}
# create base data, and pipe (%>%) to next function
create_dyadyears(directed = FALSE, subset_years = c(1816:2010)) %>%
  # subset data to politically relevant dyads (PRDs), pipe to next function
  filter_prd() %>%
  # add conflict information from GML-MID data, pipe to next function
  add_gml_mids(keep = NULL) %>%
  # add peace years
  add_peace_years()
```

Researchers should see that `{peacesciencer}`'s functionality can scale up nicely from there. For example, the following would round out the kind of information necessary to replicate Bremer's [-@bremer1992ddc] famous "dangerous dyads" analysis by adding information about national material capabilities for both states in the dyad (`add_nmc()`), estimates of democracy for both states in the dyad (`add_democracy()`), information about alliance commitments in the dyad-year (`add_cow_alliance()`), and finishing with information about estimated population size and (surplus, gross) domestic product based on simulations reported by @andersetal2020bbgb. Whereas `add_sdp_gdp()` is the last command in the pipe-based workflow, the `{peacesciencer}` call ends by assigning to an object called `Data`. This type of assignment is done with the "right hand" assignment operator (i.e. `->`).

```{r}
# create base data, and pipe (%>%) to next function
create_dyadyears(directed = FALSE, subset_years = c(1816:2010)) %>%
  # subset data to politically relevant dyads (PRDs), pipe to next function
  filter_prd() %>%
  # add conflict information from GML-MID data, pipe to next function
  add_gml_mids(keep = NULL) %>%
  # add peace years, pipe to next function
  add_peace_years() %>%
  # add capabilities data, pipe to next function
  add_nmc() %>%
  # add some estimates about democracy for each state, pipe to next function
  add_democracy() %>%
  # add information about alliance commitments in dyad-year
  add_cow_alliance() %>%
  # finish with information about population and GDP/SDP
  # and then assign to object, called, minimally, 'Data'
  add_sdp_gdp() -> Data

Data
```

If the user wants to move these data into Stata for analysis, they can save it to their current working directory with a command like `haven::write_dta(Data, "my-data.dta")` and import it into Stata when they are done. No matter, `{peacesciencer}` has pre-processed, cleaned, recoded, and merged the desired data that has greatly reduced the time and energy a researcher might otherwise spend doing something like hard-coding `-9s` in these data to be `NA` in the National Material Capabilities data. In this particular application, it has already created the main data required for a replication of @bremer1992ddc. There is only some slight data work to create the desired indicators for a statistical model of conflict onset, like a dummy variable for land-contiguity, the presence of a major power in the dyad, and some "weak-link" indicators of militarization, relative power in the dyad, level of democracy in the dyad (using the @marquez2016qme for extending the Unified Democracy Scores data), and the GDP per capita in the dyad. Table \@ref(tab:tab-dd) is a formatted version of the results of a logistic regression model of conflict onset using these "dangerous dyads" indicators. Users typically do not end their analysis here---often looking for new predictors of conflict onset with these covariates in mind---but `{peacesciencer}` greatly reduces the time and energy researchers must invest into cleaning and processing data for analysis.

```{r, echo=F, eval=F}
Data %>%
  mutate(landcontig = ifelse(conttype == 1, 1, 0)) %>%
  mutate(cowmajdyad = ifelse(cowmaj1 == 1 | cowmaj2 == 1, 1, 0)) %>%
  # Create estimate of militarization as milper/tpop
  # Then make a weak-link
  mutate(milit1 = milper1/tpop1,
         milit2 = milper2/tpop2,
         minmilit = ifelse(milit1 > milit2,
                           milit2, milit1)) %>%
  # create CINC proportion (lower over higher)
  mutate(cincprop = ifelse(cinc1 > cinc2,
                           cinc2/cinc1, cinc1/cinc2)) %>%
  # create weak-link specification using Quick UDS data
  mutate(mindemest = ifelse(xm_qudsest1 > xm_qudsest2,
                            xm_qudsest2, xm_qudsest1)) %>%
  # Create "weak-link" measure of jointly advanced economies
  mutate(minwbgdppc = ifelse(wbgdppc2011est1 > wbgdppc2011est2,
                             wbgdppc2011est2, wbgdppc2011est1)) -> Data
```

@@

## A Basic Replication of Fearon and Laitin (2003) in `{peacesciencer}`

`{peacesciencer}` is capable of generating data for replications of analyses at multiple levels beyond dyad-year, including leader-year, leader-dyad-year, and state-year. Suppose a researcher wants to create a state-year data frame to conduct an analysis of civil conflict onset analogous to Fearon and Laitin's [-@fearonlaitin2003eicw] well-cited analysis of civil conflict onset, but using UCDP conflict data and the Gleditsch-Ward state system for creating the appropriate universe of state-years. The pipe-based workflow will start with `create_stateyears(system = 'gw', subset_years = c(1946:2019))`, creating the full universe of Gleditsch-Ward state years and subsetting them to just 1946-2019 (because the UCDP data included in `{peacesciencer}` include just the observations in that time frame). Next, we can use the `add_ucdp_acd()` function to return information about ongoing UCDP conflicts and onsets for these states. `add_ucdp_acd()` takes three arguments: `type`, `issue`, and `only_wars`. `type` is an optional argument for the type of armed conflicts for which the researcher wants information. Options include "extrasystemic", "interstate", "intrastate", and "II" (short for "internationalized intrastate"). If no `type` is specified, the function returns information about ongoing disputes and onsets for all states for all tyeps of conflict. If the user wants information about multiple types of conflict--say: intra-state wars and internationalized intra-state wars---they can specify that as a character vector (e.g. `type = c("intrastate", "II")`). `issue` is another optional argument for what issue types of conflicts the user wants beyond the type of armed conflict. Options include "territory", "government", and "both". If no `issue` is specified, the function returns information for all conflicts irregarding the particular issue. `only_wars` is an argument that subsets the data to just those with the intensity levels of "war" when `only_wars = TRUE`. The argument defaults to `FALSE`, returning information about conflicts with at least 25 deaths in addition to the conflicts with more than 1,000 deaths. In this application, `add_ucdp_acd(type = "intrastate", only_wars = FALSE))` returns state-year information about ongoing intra-state conflicts over any issue and at either of UCDP's severity thresholds.[^adducdpacdalso]

@@

```{r, echo=T, eval=F}
create_stateyears(system = 'gw', subset_years = c(1946:2019)) %>%
  filter(year %in% c(1946:2019)) %>%
  add_ucdp_acd(type="intrastate", only_wars = FALSE) %>%
  add_peace_years() %>%
  add_democracy() %>%
  add_creg_fractionalization() %>%
  add_sdp_gdp() %>%
  add_rugged_terrain() -> Data
```

@@

```{r, echo=T, eval=F}
create_stateyears(system = 'gw', subset_years = c(1946:2019)) %>%
  add_ucdp_acd(type="intrastate", only_wars = TRUE) %>%
  add_peace_years() %>%
  rename_at(vars(ucdpongoing:ucdpspell), ~paste0("war_", .)) %>%
  left_join(Data, .) -> Data
```

@@

`{peacesciencer}` is already more than capable of creating the kind of data in high demand in peace science. It can create dyad-year, leader-year, leader-dyad-year, and state-year data (among others). It is also generalizable to the dispute data included in the package, allowing for merging into *dispute-year* data as well. This feature showed how it can effectively approximate two types of analyses in wide use in the peace science community. Surely researchers can and will add more information to these simple analyses after using `{peacesciencer}`, but the package already does a lot of the tedious work for researchers. It also does this in a maximally transparent way that conforms well to the DA-RT initiative across all political science. This is not to say `{peacesciencer}` does everything, but `{peacesciencer}` can only evolve and expand on what it already does well. This package can only evolve to meet new analytical demands for the peace science community. Users are free to request new features as "issues" on the project's Github.
